Chaotic dispersal of tidal debris

### 2015-09-22 ms.tex
more ref crap

In this \emph{Article}, we study the effect of chaotic diffusion of the fundamental frequencies of individual orbits on the density evolution of tidal debris. We choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics for grids of constant-energy orbits, and compare these results to measures of the density evolution of finite-volume ensembles of orbits (meant to mimic tidal debris). We find that even when the chaotic timescale is predicted to be long for a given orbit, chaos may manifest over much shorter times in small orbit ensembles through the chaotic diffusion of the individual orbits. For a chaotic orbit, the frequency spectrum of the orbit evolves with time: for a small ensemble, the spread in frequencies is therefore time-dependent, which could enhance phase-mixing. This idea supports a reevaluation of the importance of chaos in galactic haloes and implies that the amount of chaos in a given potential may have significant consequences for the observability and survivability of thin, cold tidal streams.

@@

Chaotic orbits have no orbital actions and only conserve energy (if the potential is time-independent). The orbits therefore do not have a single set of fundamental frequencies, \chchchanges{but rather the frequencies that describe the character of motion evolve with time.} In near-integrable potentials, \chchchanges{the frequencies of consecutive sub-sections of a chaotic orbit diffuse both around resonance layers\footnote{Note that during this diffusion, the frequencies never exactly hit those of a KAM torus but evolve stochastically around these discrete, stable tori \citep[cf. Figure 2 in][]{laskar99}.} (a sort of stochastic libration) and along the stochastic layers that surround the resonances (Arnold diffusion).} For weakly chaotic orbits, motion across a resonance layer can occur with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is small, motion across a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can not change by large factors (unless there is resonance overlap and the motion is strongly chaotic).

@@

We start by considering four particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular, near-resonant orbit (A), a regular, non-resonant orbit (B), a weakly chaotic orbit (C), and a \chchchanges{more} strongly chaotic orbit (D). The orbits were chosen to be close on the orbit grid so that their orbital properties (e.g., apocenter, pericenter) are similar, but have different frequency diffusion times; all four orbits are long-axis tube orbits. Figure~\ref{fig:orbits} shows the orbits in projection over \chchchanges{an integration period of 1024 orbital periods}. The initial conditions and chaos diagnostics for each orbit are listed in Table~\ref{tbl:orbit-info}. Figure~\ref{fig:ensembles} shows final positions of test-particle ensembles initialized around the four orbits described above and evolved for 64 orbital periods. A thin stream forms on the near-resonant orbit (left column), a thin---but more two-dimensional---stream forms on the non-resonant orbit (middle-left), a more diffuse stream forms on the weakly chaotic orbit with a slightly `fanned' morphology (middle-right), and a two-dimensional, `fanned' stream on the \chchchanges{more} strongly chaotic orbit (right). Given the long Lyapunov and frequency diffusion times of the weakly chaotic orbit ($t_\lambda/\periods > 900$, $t_\Omega/\periods \approx 2\times10^5$), it is surprising that the density evolution of the ensemble on this orbit appears to be more diffuse than the regular orbit stream.

@@

\item \emph{Time dependence}: Galaxies are certainly not static systems. To first order, galaxies grow in mass---for example, the spherically averaged mass profile of dark-matter haloes evolves fairly predictably in cosmological simulations \citep{wechsler02, buist14} after some initial period of stochastic mass growth. The Milky Way has probably had a fairly calm accretion history over the last 6 Gyr and therefore the mass growth may be similar to that seen in simulations. This steady growth most likely does not alter the global structure or shape of the potential. However, we also know from simulations that figure rotation, baryonic feedback, and the accretion and phase-mixing of subhaloes do perturb the global state of simulated haloes. \cite{deibel11} showed that by adopting pattern speeds comparable to those found in cosmological simulations, figure rotation generally acts to destabilize orbits (rather than stabilize chaotic orbits in the equivalent static potential) and the resulting orbit structure is that most regular orbits are associated with resonances. In future work we will explore the effect of these time-dependent processes on the chaotic dispersal of tidal streams using live potentials from cosmological N-body simulations.

@@

% Figure ??
\begin{figure}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/nbody.pdf}
\caption{ We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on the three orbits of Figure~\ref{fig:ensembles}. The progenitor in each simulation is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. We start the progenitor systems at apocenter and evolve for $\approx$64 orbital periods so that each simulation finishes again at subsequent apocenter. The mass of the progenitor is set to $10^4~\msun$, and the length-scale of the Plummer sphere is set to 5~pc to get $\approx$50--75\% mass loss over the integration time. Panels show particle positions from the final snapshots of the simulations---for visualization the positions have been rotated so that the angular momentum of the progenitor at the final snapshot are aligned with the $z$-axis. These simulations confirm that the test-particle orbit ensembles (Figure~\ref{fig:ensembles}) do (qualitatively) capture the nature of the early-stripped tidal debris. } 
\label{fig:nbodysims}
\end{center}
\end{figure}

>>>>

In this \emph{Article}, we study the effect of chaotic diffusion of the fundamental frequencies of individual orbits on the density evolution of tidal debris. We choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics for grids of constant-energy orbits, and compare these results to measures of the density evolution of finite-volume ensembles of orbits (meant to mimic tidal debris). We find that even when the chaotic timescale is predicted to be long for a given orbit, chaos may manifest over much shorter times in small orbit ensembles through the chaotic diffusion of the \chchchanges{constituent} orbits. For a chaotic orbit, the frequency spectrum of the orbit evolves with time: for a small ensemble, the spread in frequencies is therefore time-dependent, which could enhance phase-mixing. This idea supports a reevaluation of the importance of chaos in galactic haloes and implies that the amount of chaos in a given potential may have significant consequences for the observability and survivability of thin, cold tidal streams.

@@

Chaotic orbits have no orbital actions and only conserve energy (if the potential is time-independent). The orbits therefore do not have a single set of fundamental frequencies, \chchchanges{but rather the frequencies that describe the character of motion evolve with time.} In near-integrable potentials, \chchchanges{the frequencies of consecutive sub-sections of a chaotic orbit diffuse both around resonance layers\footnote{Note that during this diffusion, the frequencies never exactly hit those of a KAM torus but evolve stochastically around these discrete, stable tori \citep[cf. Figure 2 in][]{laskar99}.} (a sort of stochastic libration) and along the stochastic layers that surround the resonances (Arnold diffusion).} For weakly chaotic orbits, motion \chchchanges{around} a resonance layer can occur with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is small, motion \chchchanges{around} a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can not change by large factors (\chchchanges{unless there are other nearby resonances and overlapping stochastic layers, in which case the motion may be strongly chaotic}).

@@

We start by considering four particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular, near-resonant orbit (A), a regular, non-resonant orbit (B), a weakly chaotic orbit (C), and a \chchchanges{more} strongly chaotic orbit (D). The orbits were chosen to be close on the orbit grid so that their orbital properties (e.g., apocenter, pericenter) are similar, but have different frequency diffusion times; all four orbits are long-axis tube orbits. Figure~\ref{fig:orbits} shows the orbits in projection over \chchchanges{an integration period of 1024 orbital periods}. The initial conditions and chaos diagnostics for each orbit are listed in Table~\ref{tbl:orbit-info}. Figure~\ref{fig:ensembles} shows final positions of test-particle ensembles initialized around the four orbits described above and \chchchanges{integrated} for 64 orbital periods. A thin stream forms on the near-resonant orbit (left column), a thin---but more two-dimensional---stream forms on the non-resonant orbit (middle-left), a more diffuse stream forms on the weakly chaotic orbit with a slightly `fanned' morphology (middle-right), and a two-dimensional, `fanned' stream on the \chchchanges{more} strongly chaotic orbit (right). Given the long Lyapunov and frequency diffusion times of the weakly chaotic orbit ($t_\lambda/\periods > 900$, $t_\Omega/\periods \approx 2\times10^5$), it is surprising that the density evolution of the ensemble on this orbit appears to be more diffuse than the regular orbit stream.

@@

\item \emph{Time dependence}: Galaxies are certainly not static systems. To first order, galaxies grow in mass---for example, the spherically averaged mass profile of dark-matter haloes evolves fairly predictably in cosmological simulations \citep{wechsler02, buist14} after some initial period of stochastic mass growth. The Milky Way has probably had a fairly calm accretion history over the last 6 Gyr and therefore the mass growth may be similar to that seen in simulations. This steady growth most likely does not alter the global structure or shape of the potential. However, we also know from simulations that figure rotation, baryonic feedback, and the accretion and phase-mixing of subhaloes do perturb the global state of simulated haloes. \cite{deibel11} showed that by adopting pattern speeds comparable to those found in cosmological simulations, figure rotation generally acts to destabilize orbits (rather than stabilize chaotic orbits in the equivalent static potential) and the resulting orbit structure is that most regular orbits are associated with resonances. \chchchanges{The presence of and response to the Large and Small Magellanic Clouds may also introduce significant (time-dependent) perturbations to the global potential of the Milky Way \citep[e.g.,][]{besla10, gomez15}}. In future work we will explore the effect of these time-dependent processes on the chaotic dispersal of tidal streams using live potentials from cosmological N-body simulations.

@@

% Figure ??
\begin{figure}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/nbody.pdf}
\caption{ We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on the three orbits of Figure~\ref{fig:ensembles}. The progenitor in each simulation is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. We start the progenitor systems at apocenter and \chchchanges{integrate} for $\approx$64 orbital periods so that each simulation finishes again at subsequent apocenter. The mass of the progenitor is set to $10^4~\msun$, and the length-scale of the Plummer sphere is set to 5~pc to get $\approx$50--75\% mass loss over the integration time. Panels show particle positions from the final snapshots of the simulations---for visualization the positions have been rotated so that the angular momentum of the progenitor at the final snapshot are aligned with the $z$-axis. These simulations confirm that the test-particle orbit ensembles (Figure~\ref{fig:ensembles}) do (qualitatively) capture the nature of the early-stripped tidal debris. } 
\label{fig:nbodysims}
\end{center}
\end{figure}
